---
title: "Automated Digital Hillslope Position"
author: "Meyer Bohn"
date: "1/6/2021"
output: html_document
---
### Getting Started ###
SAGA GIS (Conrad et al., 2015) is required. Please install at https://sourceforge.net/projects/saga-gis/  
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(include = FALSE, warning = FALSE)
```

Install and load required packages  
```{r libraries}
libs <- c("stringr", "raster", "rgdal", "RSAGA", "foreign", "data.table")

#optional install
install.packages(libs)
#load
lapply(libs, require, character.only = TRUE)
```

Set up R project and working directory with your input raster (must be 3-m resoultion)
```{r inputs and directories}
#Designate the file path to your raster input file
rname <- "./INPUTRASTER.tif"
gname <- tools::file_path_sans_ext(basename(rname))

#Create outputs folder in working directory
o_folder <- dir.create(paste(getwd(), "/outputs", sep =""))
o_name <- paste(getwd(), "/outputs", sep ="")
```
Interface R with SAGA. Usually the `rsaga.env()` can find the `saga_cmd.exe`; if not, specify the path e.g. `work_env <- rsaga.env(path = "F:\Program Files (x86)\saga-5.0.0_x64")` 
```{r load SAGA cmd}
work_env <- rsaga.env()
work_env

#There are many more tools in SAGA than those used here. Use the below functions to find inspect libraries, modules, and usage.
rsaga.get.libraries(path = work_env$modules)

rsaga.get.modules(libs = "io_gdal", env = work_env)

rsaga.get.usage(lib = "io_gdal", module = "Export GeoTIFF", env = work_env)

rsaga.get.modules(libs = "ta_morphometry", env = work_env)

rsaga.get.usage(lib = "ta_morphometry", module = "Morphometric Features", env = work_env)
```
## Terrain Derivative Generation in SAGA

SAGA requires proprietary SGRD files to operate. This code converts your input raster to a `.sgrd` file and calculates the slope and profile curvature derivatives at the specified analysis scales using `morphometric features` - a method of Multi-scale quadratic Parameterisation developed by Wood (1996).  

### Warning - SAGA will use 100% of CPU and significant RAM in parallel - the size of the input Raster and your computing limitations will impact speed and feasibility of this process  
```{r SGRD and Morphometric Features}
#Create SGRD output name
sname <- paste(o_name, '/', gname, '.sgrd', sep='')
#import raster into SGRD fomrat
rsaga.import.gdal(rname, sname, env = work_env)  

#Calculate slope at 9-m analysis scale (3X3m cells)
rsaga.geoprocessor(lib = "ta_morphometry", module = "Morphometric Features", param = list(DEM = sname, SLOPE = paste(o_name, 'slp_009.sgrd', sep = '/'), SIZE = 3))

#Calculate profile curvature at 63-m analysis scale (21X3m cells)
rsaga.geoprocessor(lib = "ta_morphometry", module = "Morphometric Features", param = list(DEM = sname, PROFC = paste(o_name, 'prc_063.sgrd', sep = '/'), SIZE = 21))

```
Export the new rasters to GeoTIFF format.  
```{r export sgrd to tif - prc & slp}
#Define import export names
prc.s <- paste(o_name,  'prc_063.sgrd', sep = '/')
slp.s <- paste(o_name,  'slp_009.sgrd', sep = '/')
prc.t <- paste(o_name,  'prc_063.tif', sep = '/')
slp.t <- paste(o_name,  'slp_009.tif', sep = '/')

#export with GDAL
rsaga.geoprocessor(lib = "io_gdal", module = "Export GeoTIFF",param = list(GRIDS = prc.s, FILE = prc.t))
rsaga.geoprocessor(lib = "io_gdal", module = "Export GeoTIFF",param = list(GRIDS = slp.s, FILE = slp.t))
```
Relative elevation requires focal statistics. Besides DHP, this code can be modified to create rel at many differnt analysis scales. The case is the same for Profc and Slope.  
```{r relative elevation SAGA}
#Min grid at 135 analysis scale (45x3m cells)
rsaga.geoprocessor(lib = "statistics_grid", module = "Focal Statistics", param = list(GRID = sname, MIN = paste(o_name, 'min_135.sgrd', sep = '/'), KERNEL_TYPE= "Circle", KERNEL_RADIUS = 45))

#Max grid at 135 analysis scale (45x3m cells)
rsaga.geoprocessor(lib = "statistics_grid", module = "Focal Statistics", param = list(GRID = sname, MAX = paste(o_name, 'max_135.sgrd', sep = '/'), KERNEL_TYPE= "Circle", KERNEL_RADIUS = 45))

#import export sgrd and tif names
min.s <- paste(o_name,  'min_135.sgrd', sep = '/')
max.s <- paste(o_name,  'max_135.sgrd', sep = '/')
min.t <- paste(o_name,  'min_135.tif', sep = '/')
max.t <- paste(o_name,  'max_135.tif', sep = '/')

#export to tif like earlier
rsaga.geoprocessor(lib = "io_gdal", module = "Export GeoTIFF",param = list(GRIDS = min.s, FILE = min.t))
rsaga.geoprocessor(lib = "io_gdal", module = "Export GeoTIFF",param = list(GRIDS = max.s, FILE = max.t))
```
Once the derivatives are created and exported, we will load the inputs and create the function with `overlay()` which is think is slightly faster than the basic raster algebra.  

```{r relative elevation function}
#load original raster
ras <- raster(rname)

#load min max rasters
minras <- raster(min.t)
maxras <- raster(max.t)

#function for overlay
rel.elevation <- function(ras, minras, maxras){
  return (ras - ((minras + maxras) - ras))
}
          
rel <- overlay(ras, minras, maxras, fun=rel.elevation)
```
## Putting it all together - Classification
This portion reclassifies the terrain derivatives into the classes specified by the DHP algorithm, i.e. breaks most agreed upon by field pedologists. This would be a good place to experiment with different breaks, as the phenomenon scale of dominant pedogenic hillslope processes undoubtedly changes accordingly with different soil-forming environments.  

This function works like the Con tool in ArcGIS. 
```{r Con}
#Simple conditional for breaks
Con <- function(condition, trueValue,falseValue){
  return (condition*trueValue+(!condition)*falseValue)
}

#define rasters
slpras <- raster(slp.t)
prcras <- raster(prc.t)
relras <- rel

#2 classes 1 break - simple Con
prc_cls <- Con(prcras < 0, -1, 1)
rel_cls <- Con(relras > 0, 1, -1)

#Slope class requires 3 classes and 2 breaks - we can nest multiple Cons to achieve this
slp_cls <- Con(slpras < 1.4,-1, Con(slpras >= 1.4 & slpras <= 2.9, 0, 1))
```
Next, we create a function to classify all 5 hillslope positions.  

```{r DHP}
#Conditional statement function for DHP classification
DHPtree <- function(slpras, prcras, relras){
  Con(slp_cls == 1, 3, Con(slp_cls == 0, Con(prc_cls == 1, 2, 4), Con(rel_cls ==1, 1, 5)))
}

#Run Classification
DHP <- DHPtree(slpras, prcras, relras)
```
Finally, we need to create a factor-level attribute table for the DHP raster. Then we can export the raster to a GeoTIFF and attribute table as a csv file. 
```{r export}
#creates factors from class values and raster attribute table (RAT) - ratify
DHP <- as.factor(DHP)
DHP <- ratify(DHP, count = TRUE)

#Assigns character codes and hillslope positions
rat <- levels(DHP)[[1]]
rat$DHP_code <- (c("1","2","3","4","5"))
rat$DHP <- c("Summit", "Shoulder", "Backslope", "Footslope", "Toeslope")
levels(DHP) <- rat

#export raster to outputs folder
writeRaster(DHP, paste(o_name, "DHP.tif", sep = '/'), format = "GTiff")

#create table to join in ArcGIS
rat.csv <- as.data.table(rat)
write.csv(rat.csv,paste(o_name, "DHP.tif.vat.csv", sep = '/'), TRUE)
```
Thats it. Be sure to download the customized DHP.lyr (ArcMap) and DHP.lyrx (ArcGIS Pro) layer files from the repository for custom symbology. Stay tuned for a floodplain classificaton update (class 6) and machine-learning/big data calibration for DHP 2.0.

### "Every valley shall be raised up, every mountain and hill made low; the rough ground shall become level, the rugged places a plain" - Jesus on hillslope evolution
